# Preparing your tree and data for PCMs in R

Before we do any analysis in R we will generally need to clean and tidy our data. Data cleaning involves looking for errors, for example we might need to remove duplicate columns, or correct typos in species names, or make sure column headers are all unique and easy enough for us to type in R. We also often need to tidy our data. For data to be *tidy*:

* Each variable must have its own column.
* Each observation must have its own row.
* Each value must have its own cell.

The R packages `dplyr`, `tidyr` are great for data cleaning and tidying. See [https://r4ds.had.co.nz/tidy-data.html](https://r4ds.had.co.nz/tidy-data.html) for more details on what tidy data is and how to do this in R. Of course you are welcome to do this in Excel or another package if you're more comfortable with that. The important thing is that when you begin your analyses, your data are in a suitable format. Note that a detailed introduction to data manipulation in R is beyond the scope of this *Primer*, but if in doubt about your data, try and make it look as much like our example datasets as possible. If you can do that, these R analyses should work for you! :)

In this exercise we will take an existing tree and some data and prepare them for a PCM analysis. Note that prior to this step, it is important that your data is in a tidy format, and has been thoroughly checked for errors. The exercise below only shows you how to deal with additional issues caused by using trees and then matching species data to those trees. We assume the tree and data themselves are fit for purpose.

We will be using the evolution of eye size in frogs as an example. The data and modified tree come from Thomas et al 2020 (CITE PROPERLY), and the original tree comes from Feng et al (2019; CITE PROPERLY). I've removed a few species and a few variables to make things a bit more straightforward. If you want to see the full results check out Thomas et al 2020 (CITE PROPERLY)!

## Before you start
- Open the `04_Preparation.RProj` file in the `04_Preparation` folder to open your R Project for this exercise.
- Make yourself a new R script for your code.

You will also need to install the following packages: 

* `ape`
* `geiger`
* `tidyverse`
* `treeplyr`

***
## Preparation 
### Load the required packages
To begin we need to load the packages for this practical.

```{r, message = FALSE}
# Load packages
library(ape)
library(geiger)
library(tidyverse)
library(treeplyr)
```

## Reading and checking your phylogeny in R

We already learned how to do these things in `03-Phylogenies.Rmd`. If you haven't looked at that exercise I suggest you check it out before doing the steps below.

First let's read in the tree and look at its structure:
```{r}
# Read in the tree
frogtree <- read.nexus("data/frog-tree.nex")
# Look at the tree summary
frogtree
```

It's usually a good idea to quickly plot the tree too...
```{r}
# Plot the tree as a circular/fan phylogeny with small labels
plot(frogtree, cex = 0.2, typ = "fan", no.margin = TRUE)
```

`frogtree` is a fully resolved tree with branch lengths. There are 214 species and 213 internal nodes. Note that the species names at the tips also have family names added to them...

```{r}
# Look at the first 5 tip labels
frogtree$tip.label[1:5]
```
Most trees will have just the genus and species names, but some will have additional information like here, or numbering etc. This is not a problem as long as the names match those in your data (see below).

Most R functions require your tree to be dichotomous, i.e. to have no polytomies. To check whether your tree is dichotomous use `is.binary.tree`. 

```{r}
# Check whether the tree is binary
# We want this to be TRUE
is.binary.tree(frogtree) 
```

If this was FALSE, we'd use `multi2di` to make the tree dichotomous, but here it is TRUE so we can leave it as it is.

Most functions also require the tree to be rooted, i.e., to have one taxon designated as the outgroup. We can check whether the tree is rooted as follows.

```{r}
# Check whether the tree is rooted
# We want this to be TRUE
is.rooted(frogtree)
```

Our tree is rooted so it's ready to go.

## Reading the data into R
The data are in a comma-delimited text file called `frog-eyes.csv`. Load these data as follows.

```{r}
# Read in the data
frogdata <- read_csv("data/frog-eyes.csv")
```

Check everything loaded correctly:

```{r}
# Look at the data
glimpse(frogdata)
```

As you can see, the data contains 215 species, and the following 11 variables: 

* `Binomial` - the species binomial name.        
* `Family` - the family the species belongs to.        
* `Genus` - the genus the species belongs to.              
* `tiplabel` - the name used for the species in the phylogeny.       
* `Adult_habitat` - habitat of adults. Categories are: Ground-dwelling, Subfossorial, Scansorial (i.e. tree-dwelling), Semiaquatic, Aquatic, or Fossorial (i.e. burrowers).       
* `Life_history` - whether the larvae are free-living (Free-living larvae) or not (No free-living larvae).     
* `Sex_dichromatism` - whether different sexes are different colours (Present) or not (Absent).
* `SVL` - snout vent length (in mm). This is a common way to measure body size in amphibians.            
* `mass` - body mass (in g).
* `rootmass` - cube root of the body mass.        
* `eyesize` - eye size (in mm) for the species. This is an everage across left and right eyes from three individuals per species.

## Matching your data to your phylogeny
Now we have the tree and the data in R, we need to match the two up if we want to perform any kind of PCM analyses. Below are some common issues you might encounter, and how to fix them.

### Species names with spaces
Species names in phylogenies are generally written as *Genus_species* (the gap between the genus name and species name is replaced by an underscore `_`). If the species names in the data are written as *Genus species* with a space, then you will have to replace the spaces with `_` so that they match up with the species names in the tree. You can do this as follows using `str_replace`.

We don't need to do this in our frog data, but if we did we could use code like this:
```{r, eval = FALSE}
# Replace spaces with underscores in species names
frogdata <- 
  frogdata %>%
  mutate(Binomial = str_replace(string = Binomial, pattern = " ", replacement = "_"))

# Check it worked
glimpse(frogdata)
```

### Mismatches between species in your data and phylogeny
Often you will have data for species which are not in your phylogeny and/or species in your phylogeny which are not in your data. Many functions in R can deal with this and will match the species for you, others will produce an error telling you the tree and data do not match (e.g. most `ape` functions). 

Even in functions that can cope with this, it's useful to match up the species *before* your analyses. This can help you identify things like spelling mistakes or variations in the taxonomy of the tree and the data. 

```{block, type = "info"}
If you have even slightly misspelled a species name in the tree or the data it will automatically be dropped from the analyses. It is therefore **very important** to check this before running an analysis, especially one with lots of taxa.
```

We can use the `geiger` function `name.check` to find out which names do not match. Remember that the species names that match up with the tree fror `frogdata` are in the variable `tiplabel`.

```{r, message = FALSE}
# Check whether the names match in the data and the tree
check <- name.check(phy = frogtree, data = frogdata, 
                    data.names = frogdata$tiplabel)
```

The output of `check` has two parts, `tree_not_data` for species in the tree but not in the dataset, and `data_not_tree` for species in the dataset but not in the tree. You need to look at both of these in turn.

```{r}
# Look at species in the tree but not the data
check$tree_not_data
```

There are four species in the tree and not the data. We were expecting this here, so no worries.

For your analyses you should *always* check this list carefully. If I were running this analysis for the first time I'd want to check that these species really weren't in my data. Maybe they are misspelled in the data? If so correct this now. Maybe the species name has changed? If so change this now. It doesn't really matter whether you make the change in the tree or the data, but make sure anything that *should* match up, does. I'd usually fix issues in the dataset, unless it's a blatant typo in the tree.

Next check the species in the data but not the tree.

```{r}
# Look at species in the data but not the tree
check$data_not_tree
```
This gives us five species in the data but not in the tree. As above, make sure to correct any errors before moving to the next step.

Here we know that these species are missing from our data, so we don't need to worry.

### Matching the tree and the data 
finally, once we know which species do not match up, we need to remove species missing from the data from the tree, and remove species missing from the tree from the data. This used to be a bit of a pain, but `treeplyr` makes it easy. `treeplyr` has a lot of really cool functions, see the [wiki](https://github.com/uyedaj/treeplyr/wiki) for more details. However, here we are just going to use it just to match up the tree and the data.

We'll use the function `make.treedata` to combine the tree and the dataset into one object. We need to provide the name of the tree, the name of the data, and specify which column our species names are in. 

```{r}
# Combine and match the tree and data
frogstuff <- make.treedata(tree = frogtree,  data = frogdata, 
                              name_column = "tiplabel")
```

Note that we could leave out the `name_column = tiplabel` argument, in which case the function `make.treedata` will search the data for the column with the contents that have most matches to the tree, and automatically use this column for matching up species names. It will also search the rownames.

To look at the tree and data combined we can use `summary`:

```{r}
summary(frogstuff)
```

You may notice that `make.treedata` has two objects within it, the tree and the data. You can access the tree using `frogstuff$phy` and the data using `frogstuff$dat`. 

Let's look at the phylogeny first...

```{r}
# Look at the tree summary
frogstuff$phy
```

The matched phylogeny has 210 species in it (instead of the original 214 species in `frogtree` because the four that weren't in the data have been removed).

Now let's look at the data. What is missing?
```{r}
# Look at the data
glimpse(frogstuff$dat)
```

We now have 210 species in the dataset too, great! But hopefully you noticed that the column with the species names in it (`tiplabel`) has disappeared! `treeplyr` is designed to work with R functions that assume species names will be in the rownames of your data. However, not all PCMs in R work this way, and I personally like to be able to see species names when I quickly look at a dataset. We can make a new species names column using the tip labels from the phylogeny, as `make.treedata` orders the data so it's the same as the tip labels.

```{r}
# Make a new column called Species with the tip labels in it
frogstuff$dat$Species <- frogstuff$phy$tip.label
```

Finally we might want to rename the tree and data to make them a bit less clunky when typing in your code.

```{r}
# Save tree as mytree
mytree <- frogstuff$phy
# Save data as mydata
mydata <- frogstuff$dat
```

For the data, I'm also going to add one last trick to make our lives easier. `treeplyr` relies on the `tidyverse` set of packages, which create special dataframes called tibbles. Tibbles are great, *but* some older PCM functions cannot work with them, and need the data to be in an ordinary dataframe instead. We can fix this using `as.data.frame`.

```{r}
# Look at the structure of mydata
str(mydata)
```

```{r}
# Force mydata to be a data frame
mydata <- as.data.frame(mydata)
# Check mydata is now a dataframe
str(mydata)
```

Finally, we might want to output these cleaned and tidied data and tree to our folder so rather than doing this everytime we start an analysis, we can just use these tidy versions. To do this we can use:

```{r, eval = FALSE}
# Write the cleaned data to a new file
write_csv(mydata, path = "data/clean-frog-data.csv")

# Write the cleaned tree to a new file
write.nexus(mytree, file = "data/clean-frog-tree.nex")
```

Note, however, that you will have to repeat this preparation process if you add species or data to your tree or dataset.

### Subsetting your tree and data 
Another thing that `treeplyr` makes a lot easier is subsetting your tree and data. It's fairly common, especially with large phylogenies, that we might want to run our analyses on subsets of the data. Generally these are taxonomic divisions, but you might also want to divide your analyses into large and small body size species, for example.

One solution would be to make a new dataset, and then run through the same procedure as we've used above. However, we can instead just subset the tree data object itself using `filter`. This works the same as `filter` normally does in `dplyr`. As an example, let's select only species in the family Bufonidae (toads).

```{r}
# Subset only the species in the Bufonidae family 
bufonidae <- filter(frogstuff, Family == "Bufonidae")

# Plot tree to check it worked
plot(bufonidae$phy)
```

If you want to use this in later analyses, you might want to save these as separate data and phylogeny objects, and don't forget to make the data into a dataframe. 

Another thing that could be useful here is to subset so that you have a complete set of variables for certain analyses. For example let's see how many "NAs" there are in each of our variables (don't worry too much about this code if it seems confusing to you!):

```{r}
mydata %>%
  select(everything()) %>%
  summarise_all(list(~ sum(is.na(.))))
```
22 species don't have a value for `Life_history`, and 36 species don't have a value for `Sex_dichromatism`. Many PCMs will just ignore NA values, but if you needed to remove these you could subset them out as follows, leaving only the 163 species with complete data.

```{r}
# Subset out the species with NA values for some variables
frog_noNA <- filter(frogstuff, 
                    !is.na(Life_history) & !is.na(Sex_dichromatism))

# Look at the data
glimpse(frog_noNA$dat)
```

## Quick template code
To help you do this with your own data, I've condensed the above into one script below so you don't forget any of the steps.

```{r, eval = FALSE}
# Load packages
library(ape)
library(geiger)
library(tidyverse)
library(treeplyr)

# Read in the tree
frogtree <- read.nexus("data/frog-tree.nex")
# Look at the tree summary
str(frogtree)

# Plot the tree as a circular/fan phylogeny with small labels
plot(frogtree, cex = 0.2, typ = "fan", no.margin = TRUE)

# Check whether the tree is binary
# We want this to be TRUE
is.binary.tree(frogtree) 

# Check whether the tree is rooted
# We want this to be TRUE
is.rooted(frogtree)

# Read in the data
frogdata <- read_csv("data/frog-eyes.csv")

# Look at the data
glimpse(frogdata)

# Check whether the names match in the data and the tree
check <- name.check(phy = frogtree, data = frogdata, 
                    data.names = frogdata$tiplabel)
# Look at check
check

### Correct any typos/taxonomic errors in the tree or data ###
  
# Combine and match the tree and data
frogstuff <- make.treedata(tree = frogtree,  data = frogdata, 
                              name_column = "tiplabel")

# Look at the tree summary
frogstuff$phy

# Look at the data
glimpse(frogstuff$dat)

# Make a new column called Binomial with the tip labels in it
frogstuff$dat$Binomial <- frogstuff$phy$tip.label

# Save tree as mytree
mytree <- frogstuff$phy
# Save data as mydata
mydata <- frogstuff$dat

# Force mydata to be a data frame
mydata <- as.data.frame(mydata)
# Check mydata is now a dataframe
str(mydata)

## OPTIONAL
# Make any required subsets of the tree/data

## OPTIONAL
# Write the cleaned data to a new file
write_csv(mydata, path = "data/clean-frog-data.csv")

# Write the cleaned tree to a new file
write.nexus(mytree, file = "data/clean-frog-tree.nex")
```

## Practical exercises

In the data folder there is another tree (`consensusTree_10kTrees_Version2.nex`) and dataset (`primate-life-history-data.csv`) for investigating the evolution of primate life-history variables. These data come from the PanTHERIA database (Jones et al. 2009) and 10kTrees (Arnold et al. 2010). 

Read in the tree and data, then prepare them for a PCM analysis.
