# Answers to practical exercises

Below are all the questions and answers to the practical exercises at the end of each exercise. 

## Data Wrangling

Using the `frog-eyes-raw-data.csv` data and R, create a new dataframe that contains only the columns `genus_species`, `Family`, and two new columns containing the *species mean values* for `rootmass` and `eyemean`. Note that we will talk more about this dataset in later exercises, but it comes from (Thomas et al 2020 ADD CITATION)

There are multiple ways of doing this, but the easiest is as follows:
```{r}
# Load packages
library(tidyverse)

# Read in the data
frog <- read_csv("data/frog-eyes-raw-data.csv")
# Take a look at it
glimpse(frog)

# Create a new dataframe called frog2
frog2 <-
  # Start with the frog dataframe
  frog %>%
  # Select only the columns of interest
  select(genus_species, Family, rootmass, eyemean) %>%
  # Group by species name, and Family (or summarise will try and average this too!)
  group_by(genus_species, Family) %>%
  # Summarise to get the mean values
  summarise(mean_rootmass = mean(rootmass),
            mean_eyemean = mean(eyemean))

# Look at frog2
glimpse(frog2)

# Write to file
# This is commented out as I don't want to save this
# write_csv(frog2, path = "data/clean-frog-data.csv")
```

## Phylogenies

Using the `frog-tree.nex` phylogeny, see if you can do the following. Note that we will talk more about this tree in later exercises, but it comes Feng et al. (2019), and was modified in Thomas et al 2020 ADD CITATION).

1. Read the tree into R.
```{r}
# Load packages
library(ape)
library(ggtree)
library(tidyverse)

# Read in the tree
tree <- read.nexus("data/frog-tree.nex")

# Look at it
tree
```
2. Use R functions to determine: 
    a) how many species are in the tree?
    
```{r}
# Look at the tree structure
str(tree)
# Or the whole tree
tree

# There are 214 tips
```
    b) is the tree fully resolved?
```{r}
# Check whether the tree is binary/fully resolved
is.binary.tree(tree)
# Yes it is!
```
    c) is the tree rooted?
```{r}
# Check whether the tree is rooted
is.rooted(tree)
# Yes it is!
```    
3. Use `plot.phylo` to plot the tree.
    a) Can you change the size of the tip labels?
    b) Can you make a fan-shaped plot? 
    c) Can you change the colour of the tips and branches?
```{r}
# Plot the tree
plot(tree, 
     no.margin = TRUE, # to make sure it fills the whole space
     type = "fan", # to make a fan tree
     cex = 0.6, # to reduce the tip label size (default = 1)
     tip.color = "deeppink", # to change the tip colour
     edge.color = "springgreen") # to change the branch colour
# yuk!
```
4. Save the tree to file as "mysuperdoopertree.nex"    
```{r}
# Save the tree
# I've commented this out as I don't want to save it here.
# write.nexus(tree, file = "mysuperdoopertree.nex")
```    

EXTRA. Use ggtree and the ggtree manual to produce a phylogeny plot of this tree that you could use in a publication OR go make the most exotic looking tree you possibly can. Playing around with these packages is the best way to learn. If you make a particularly amazing or horrific tree, send us a screenshot!

## Preparations

In the data folder there is another tree (`consensusTree_10kTrees_Version2.nex`) and dataset (`primate-life-history-data.csv`) for investigating the evolution of primate life-history variables. These data come from the PanTHERIA database (Jones et al. 2009) and 10kTrees (Arnold et al. 2010). 

For practice, read in the tree and data then prepare them for a PCM analysis.

```{r, eval = FALSE}
# Load packages
library(ape)
library(geiger)
library(tidyverse)
library(treeplyr)

# Read in the tree
primatetree <- read.nexus("data/consensusTree_10kTrees_Version2.nex")
# Look at the tree summary
str(primatetree)

# Plot the tree as a circular/fan phylogeny with small labels
plot(primatetree, cex = 0.2, typ = "fan", no.margin = TRUE)

# Check whether the tree is binary
# We want this to be TRUE
is.binary.tree(primatetree) 

# It is FALSE so make the tree binary:
primatetree <- multi2di(primatetree)

# Check whether the tree is binary now...
is.binary.tree(primatetree) 

# Check whether the tree is rooted
# We want this to be TRUE
is.rooted(primatetree)

# Read in the data
primatedata <- read_csv("data/primate-life-history-data.csv")

# Look at the data
glimpse(primatedata)

# Replace spaces with underscores in species names
primatedata <- 
  primatedata %>%
  mutate(Binomial = str_replace(string = Binomial, pattern = " ", replacement = "_"))
# Check it worked
glimpse(primatedata)

# Check whether the names match in the data and the tree
check <- name.check(phy = primatetree, data = primatedata, 
                    data.names = primatedata$Binomial)
# Look at check
check

# 149 species are in the tree but not the data
# 0 species are in the data but not the tree
  
# Combine and match the tree and data
primatestuff <- make.treedata(tree = primatetree,  data = primatedata, 
                              name_column = "Binomial")

# Look at the tree summary
primatestuff$phy
# Look at the data
glimpse(primatestuff$dat)
## There are 77 species in the tree and the data

# Make a new column called Binomial with the tip labels in it
primatestuff$dat$Binomial <- primatestuff$phy$tip.label

# Save tree as mytree
mytree <- primatestuff$phy
# Save data as mydata
mydata <- primatestuff$dat

# Force mydata to be a data frame
mydata <- as.data.frame(mydata)
# Check mydata is now a dataframe
str(mydata)

## OPTIONAL
# Make any required subsets of the tree/data

## OPTIONAL
# Write the cleaned data to a new file
write_csv(mydata, path = "data/clean-primate-data.csv")

# Write the cleaned tree to a new file
write.nexus(mytree, file = "data/clean-primate-tree.nex")
```

## PGLS

In the data folder there is another tree (`consensusTree_10kTrees_Version2.nex`) and dataset (`primate-life-history-data.csv`) for investigating the evolution of primate life-history variables. These data come from the PanTHERIA database (Jones et al. 2009) and 10kTrees (Arnold et al. 2010). 

Let's investigate the relationship between gestation length in Primates and their body size. Theory leads us to believe that bigger mammals have longer gestation lengths (compare mice to elephants), but is this true in Primates?

Read in the tree and data, then prepare them for a PCM analysis (you should have already done this in the previous exercise which should save you some time). Then fit a PGLS model to investigate the relationship between log gestation length (y = log(GestationLen_d)) and log body size (x = log(AdultBodyMass_g)) in primates. Don't forget to look at the model diagnostics!!!

```{r, message = FALSE}
# Load the packages
library(ape)
library(geiger)
library(treeplyr)
library(caper)
library(tidyverse)

### Prepare the tree and the data as shown in the answer above ###
### This should mean you now have mydata and mytree ready to work with ###

# Prepare the data for use in caper
primate <- comparative.data(phy = mytree, data = mydata, 
                            names.col = Species, vcv = TRUE, 
                            na.omit = FALSE, warn.dropped = TRUE)

# Check the list of dropped species is what you expected
primate$dropped$tips
primate$dropped$unmatched.rows

# Fit a PGLS model
model.pgls <- pgls(log(GestationLen_d) ~ log(AdultBodyMass_g), 
                   data = frog, lambda = "ML")

# Look at the model outputs
summary(model.pgls)
### Lambda should be equal to 0.892

# Plot the results
ggplot(frogdata, aes(x = log(AdultBodyMass_g), 
                     y = log(GestationLen_d))) +
  geom_point() +
  geom_abline(slope = coefficients(model.pgls)[2], intercept = coefficients(model.pgls)[1]) +
  theme_bw()

### Model diagnostics

# Make a plotting window with 4 panes as there are four plots
par(mfrow = c(2, 2))
# Now plot the model diagnostics
plot(model.pgls)
# Return the plot window to one pane for later plotting
par(mfrow = c(1, 1))

# Create a likelihood profile of the lambda estimate
lambda.profile <- pgls.profile(model.pgls, "lambda")
# Plot the likelihood profile
plot(lambda.profile)

# Extract the confidence intervals on lambda
pgls.confint(model.pgls, "lambda")$ci.val
```